\textbf{CPSC 504 Project\\\vspace{3 cm}\Huge{An investigation into an application to manage data transformations and updates}\vspace{2 cm}}

### 2015-04-14 Final_Paper/Paper.tex
Paper.

\documentclass[12pt]{article}
\usepackage{color}
\usepackage[margin=1.0in]{geometry}

@@

\noindent Our primary contributions are:
\begin{enumerate}
	\item{Defined a data transformation language using logical programming concepts to capture changes made to a relational data model}
	\item{Design decisions about how the relational model should react and handle schema changes}
	\item{Preliminary investigation into view maintenance policies for the remote datastore}
	\item{Further detailing the scenario in which an application views a subset of columns from one or more tables in a remote dataset}
	\item{Decided which issues to assign to the server side and which to the client side}
\end{enumerate}

\section{Motivation}
In our project, our main focus is on a scenario where the user views any subset of columns from one or more tables, it is important to figure out how this relates to the real world. The example in \cite{arniThesis} talks about a sports reporter who retrieves select information about basketball players to do an article on. In this case, the reporter only takes out columns regarding a player's name, his high school, and the first team he played on after entering the NBA. Compared to all the information that is stored about NBA players on a main database (e.g., salary, weight, height, number of transfers, agent name, etc.), the reporter has only taken a very small subset of columns. Another example could be a school administrator who wants to decide which students to give awards to purely based on academic merit. He/she could go to the main database storing all student information (e.g., the database could have information regarding a student's name, student number, GPA, guardian contact information, date of birth, allergy/medication information, family doctor contact, etc.) and only choose to view the student number and GPA column so that he/she will not be biased towards any names or gender.

@@

\subsection{Sample data used in our project}
In our project, we decided to borrow the dataset used in CPSC 304, an introductory relational database course at UBC, to use as the test dataset in our application. This was a movie star dataset that had the columns ``Star Id'', ``Name'', and ``Gender''. After supplementing some additional data to the dataset, namely the columns ``Movie Title'', ``Production Studio'', and ``Critic Rating'', we used this dataset. We also added in a second table detailing user information with the columns ``Movie Title'' which is a foreign key referencing the ``Movie Title'' column in the Movie Information table, ``Website that user information was obtained from'', ``User Rating'', and ``Number of Likes''.

\section{Solution}
The basic workflow of the scenario described in Section \ref{sec:introduction} will remain the same. However, as the user obtains more flexibility over what columns can be chosen to view in his/her application,  the complexity of the data transformations increase along with the amount of information that must be tracked (see Section \ref{sec:transformation_language}). In this project, we have decided to support the following user changes: adding a column, removing a column, editing a column name, changing a table cell's current value, removing a table cell's value to make it null, adding a row, and removing a row. Any changes that cause a violation in schema definitions (e.g., removing a row's primary key) will be resolved on the server side by the database administrator. This design decision was due to not wanting to restrict the types of users that would be able to use an app to find data to examine. As we have to assume that there could be users unfamiliar with schema/storage restrictions, we decided to ensure that schema violations are resolved by someone with the technical understanding of what error has occurred.

@@

\section{Related Work}
\textcolor{red}{need stuff here can probably reuse the stuff from the project update but will need more + editing} \\
\textbf{\textcolor{green}{Stuff in this section is ripped from the project update}}

\subsection{ETL: Extract, Transform, Load}

\subsection{Transformation Languages}
While there has been a large body of work done on various data transformation tools \cite{kandel2011wrangler, raman2001potter, kandel2012profiler}, these tools often focus on visualizing the data transformation rather than the transformation language used to represent the data changes behind the scenes. In fact, most of the work in the data transformation language domain seems to be classified into one of two categories. Either the work focuses on trying to help uses visualize the data transformations that have occurred \cite{kandel2011wrangler, raman2001potter, kandel2012profiler} or it focuses on modifying existing transformation languages to help with the data transformation process \cite{ilprints409, lakshmanan2001schemasql}. Due to the lack of literature about the specifics of data transformation languages, we decided to cast our net further and examine data transformation language frameworks. There was a minimal amount of literature on using logical programming concepts to create a data transformation language \textcolor{red}{need citation}. The idea was that people find declarative languages like logical programming languages easier to understand as it focuses on the relationships between the data models rather than the method of how to locate a certain piece of information to work with \textcolor{red}{need citation}. There has been research working with declarative data transformation languages for different data models \cite{bry2002towards, lawley2006practical, tarau2009embedded} so we decided to try and model our data transformation language accordingly.

\textcolor{red}{incorporate into paragraph}
\cite{bry2002towards} : declarative data transformation language for XML and semi-structured data

\cite{lawley2006practical} : implementation of a pattern based language for transforming data models

\cite{tarau2009embedded} : embedded logical language framework

@@

\subsection{View Maintenance}
Currently we have chosen to only allow data from a single table be viewed in our interface. However, given that this project could be extended in the future to include viewing data that is from multiple tables, we are looking ahead at literature in the view maintenance domain to try and predict what challenges our interface might face \cite{ agrawal1997efficient, agrawal2009asynchronous, gupta1999materialized, zhou2007lazy}. We also thought that we might be able to get some ideas on how often to push any possible schema mapping/schema changes as it seems to be a comparable problem to how materialized views need to push its changes to a remote database \cite{agrawal1997efficient, agrawal2009asynchronous, zhou2007lazy}.

@@

>>>>

\documentclass[12pt]{article}
\usepackage{color}
\usepackage[margin=1.0in]{geometry}
\newcommand{\hilight}[1]{\colorbox{yellow}{#1}}

@@

\noindent \textcolor{blue}{do we want to put the examples here?}

\noindent Our primary contributions are:
\begin{enumerate}
	\item{Defined a data transformation language using logical programming concepts to capture changes made to a relational data model}
	\item{Proposed how the relational model could react and handle schema changes}
	\item{Investigated view maintenance policies for the remote datastore}
	\item{Detailed scenarios in which an application views a subset of columns from one or more tables in a remote dataset}
	\item{Decided which issues to assign to the server side and which to the client side}
\end{enumerate}

\section{Motivation}
In our project, our main focus is on a scenario where the user views any subset of columns from one or more tables. It is important to figure out how this relates to the real world.  For instance, a sports reporter wishes to write an article on a select number of basketball players and retrieves a specific set of information related to the chosen players \cite{arniThesis}. In this case, the reporter only uses columns such as player's name, his high school, and the first team he played for after entering the NBA. Compared to all the information that is stored about NBA players on a main database (e.g., salary, weight, height, number of transfers, agent name, etc.), the reporter has only looked a small subset of columns.

Consider another example: at the end of a school year, a school administrator is deciding the students to whom academic merit prizes are to be awarded.  He/she could go to the main database storing all student information (e.g., the database stores detailed student information including a name, student number, GPA, guardian contact information, date of birth, allergy/medication information, family doctor contact, etc.) but only wishes to access student number and GPA columns to ensure that he/she will not be biased towards any name or gender.

@@

\subsection{Sample data used in our project}
In our project, we decided to borrow the dataset used in CPSC 304, an introductory relational database course at UBC, to use as the test dataset in our application. \hilight{This was a movie star dataset that had the columns ``Star Id'', ``Name'', and ``Gender''. After supplementing some additional data to the dataset, namely the columns ``Movie Title'', ``Production Studio'', and ``Critic Rating'', we used this dataset.} \textcolor{blue} {awkward wording} We also added in a second table detailing user information with the columns ``Movie Title'' which is a foreign key referencing the ``Movie Title'' column in the Movie Information table, ``Website that user information was obtained from'', ``User Rating'', and ``Number of Likes''.

\section{Solution}
The basic workflow of the scenario described in Section \ref{sec:introduction} remains the same. However, as the user obtains more flexibility over what columns can be chosen to view in his/her application,  the complexity of the data transformations increase along with the amount of information that must be tracked (see Section \ref{sec:transformation_language}). In this project, we have decided to support the following user changes: adding a column, removing a column, editing a column name, changing a table cell's current value, removing a table cell's value to make it null, adding a row, and removing a row. Any changes that cause a violation in schema definitions (e.g., removing a row's primary key) will be resolved on the server side by the database administrator. This design decision was due to not wanting to restrict the types of users that would be able to use an app to find data to examine. As we have to assume that there could be users unfamiliar with schema/storage restrictions, we decided to ensure that schema violations are resolved by someone with the technical understanding of what error has occurred.

@@

\section{Related Work}
\subsection{Transformation Languages}
While there has been a large body of work done on various data transformation tools \cite{kandel2011wrangler, raman2001potter, kandel2012profiler}, these tools often focus on visualizing the data transformation rather than the transformation language used to represent the data changes behind the scenes. In fact, most of the work in the data transformation language domain seems to be classified into one of two categories. Either the work focuses on trying to help uses visualize the data transformations that have occurred \cite{kandel2011wrangler, raman2001potter, kandel2012profiler} or it focuses on modifying existing transformation languages to help with the data transformation process \cite{ilprints409, lakshmanan2001schemasql}. Due to the lack of literature about the specifics of data transformation languages, we decided to cast our net further and examine data transformation language frameworks. There was a minimal amount of literature on using logical programming concepts to create a data transformation language \cite{bry2002towards}. The idea was that people find declarative languages like logical programming languages easier to understand as it focuses on the relationships between the data models rather than the method of how to locate a certain piece of information to work with \cite{bry2002towards, tarau2009embedded}. There has been research working with declarative data transformation languages for different data models, highlighting the flexibility requirements in order to keep current with today's less structured data, \cite{bry2002towards, lawley2006practical, tarau2009embedded} so we decided to try and model our data transformation language accordingly.

@@

\subsection{View Maintenance}
As our application takes a mixed-initiative approach to updating the remote data source, we also thought that we might be able to get some ideas on how often to push any possible schema mapping/schema changes as it seems to be a comparable problem to how materialized views need to push its changes to a remote database \cite{agrawal1997efficient, agrawal2009asynchronous, zhou2007lazy}. Through looking at how views and indices are maintained in a large distributed database \cite{agrawal1997efficient}, we realized that many of the design decisions we made for our transformation language (see Section \ref{sec:transformation_language}) fit quite well with the model suggested in the paper. For example, their discussion on the different ways to push updates and the decisions made on whether the system or the application should handle certain update issues do mirror our decisions.

@@

The update policy of this system can be examined more closely to see if we can draw upon more of the principles and ideas discussed in the distributed database-application system PNUTS \cite{agrawal1997efficient}.
